Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
do not hyphenate hidden Markov model

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden-Markov model has a
\emph{match state}.
\remark{Time to introduce the elements of Viterbi's equations!}
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

@@

This directed graph represents the states of a
hidden-Markov model in the ``Plan7'' format defined by the HMMER
software package.
%% Only 7 of the possible 9 state transitions are allowed, which 
%% simplifies the dynamic programming.
Each insertion state~($I$) represents a potential mutation that 
inserts an amino acid;
each deletion state~($D$) represents a potental mutation that
deletes an amino acid;
and each match state~($M$) represents either
an unmutated amino acid or a potential mutation that
substitutes one amino acid for another.
A~match state or an insertion state includes a table that assigns a
nonzero probability to each amino acid.
(The figure shows tables for the first two states only.)
A~triplet of an insertion state, deletion state, 
and match state formas a \emph{node}, which corresponds to a column in
the alignment used to train the model.
(Each node is shown in a dashed oval.)

\caption{A hidden-Markov model trained from a four-column alignment}

@@

A hidden-Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden-Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967}.

>>>>

The protein alignment is used to train a \emph{hidden Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
\remark{Time to introduce the elements of Viterbi's equations!}
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

@@

This directed graph represents the states of a
hidden Markov model in the ``Plan7'' format defined by the HMMER
software package.
%% Only 7 of the possible 9 state transitions are allowed, which 
%% simplifies the dynamic programming.
Each insertion state~($I$) represents a potential mutation that 
inserts an amino acid;
each deletion state~($D$) represents a potental mutation that
deletes an amino acid;
and each match state~($M$) represents either
an unmutated amino acid or a potential mutation that
substitutes one amino acid for another.
A~match state or an insertion state includes a table that assigns a
nonzero probability to each amino acid.
(The figure shows tables for the first two states only.)
A~triplet of an insertion state, deletion state, 
and match state formas a \emph{node}, which corresponds to a column in
the alignment used to train the model.
(Each node is shown in a dashed oval.)

\caption{A hidden Markov model trained from a four-column alignment}

@@

A hidden Markov model like that shown in \figref{plan7} can be used to
determine the likelihood that a new protein shares structure and
evolutionary history with the proteins used to train the model.
(The new protein's amino-acid sequence is known, but its structure and
function are unknown.)
Such a likelihood can be computed by
a~widely used software package called HMMER (``hammer'').
HMMER computes the optimal \textit{parse}, or 
alignment, of a protein sequence onto the directed graph of the hidden Markov 
model.
A parse maps each amino acid of the sequence under test onto a state
of the model; equivalently, a parse is a \textit{path} through the model.
Each edge is weighted with a probability, and each emitting 
state (match or insertion) maps each amino acid to a probability,
and the 
probability of a given path is the product of these individual probabilities.
The most likely path, and its associated probability, are computed 
using an algorithm developed by~\citet{Viterbi:1967}.
